---
title: "Data Wrangling Project"
author: "Kouame J Oura"
date: "4/10/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, message=FALSE, warning=FALSE}
library(jsonlite)
library(lubridate)
library(dplyr)
library(gridExtra)
library(ggplot2)
library(magrittr)
library(stringr)
library(purrr)
library(tidyr)
library(plotly)
library(tidyverse)
```

#                               ###### NEW YORK CITY MOTOR VEHICLE COLLISION #######

The Motor Vehicle Collisions-Crashes* dataset from NYC Open Data will be used. It is provided by the Police Department(NYPD). The dataset has 1.77 million rows and 29 columns, each row contains details about crash event in New York city. It is downloaded using R programming, API Endpoint and saved to the working directory as a .json file. The goal of this project is to use R tools to explore and analyze New York City Motor Vehicle collisions.

## Data

##### Downloading data once and save it locally
This is to avoid downloading the data multiple times, saving time.

The data has 1768292 rows and 29 variables

```{r}
#### https://dmv.ny.gov/forms/p33.pdf Police Crash Report Submission Instructions
crash_url = "https://data.cityofnewyork.us/resource/h9gi-nx95.json?$limit=2000000"
GET(crash_url, write_disk(path="crash.json", overwrite=TRUE))
```

## Data cleaning and tranformation

Loading only needed variables

```{r}
crash_data = jsonlite::fromJSON("crash.json")
crash_data =crash_data[ , c("crash_date", "crash_time", "borough", "zip_code", "latitude", "longitude", "number_of_persons_injured", "number_of_persons_killed", "number_of_pedestrians_injured", "number_of_pedestrians_killed", "number_of_cyclist_injured", "number_of_cyclist_killed", "number_of_motorist_injured", "number_of_motorist_killed", "contributing_factor_vehicle_1", "contributing_factor_vehicle_2", "collision_id", "vehicle_type_code1", "vehicle_type_code2")]
```

#### First six rows of data

To have an idea of the data
```{r}
crash_data = as_tibble(crash_data) # convert data.frame to tibble
crash_data
```
#### Records missing accurate location
Identifying records missing longitude or latitude.
```{r}
na_loc=sum(is.na(as.numeric(crash_data$longitude) & as.numeric(crash_data$latitude)))
na_loc # records missing longitude or latitude
```

Removing records missing location longitude or latitude coordinate

```{r}
crash_data = crash_data[! is.na(crash_data$longitude),]
crash_data = crash_data[! is.na(crash_data$latitude),]
```

#### Primary key check
Checking if there is a duplicate record

```{r message=FALSE, warning=FALSE}
library(sets)
num_dup=sum(is.na(crash_data$collision_id)) # Checking for missing primary key
anyDuplicated(crash_data$collision_id) # Checking for duplicate records
num_dup
```

#### Crash time format

Removing the trailing component "T00:00:00.000" from crash_date,
then rename concatenation of crash_date and crash_time as
crash_time in the ymd_hm date format

```{r}
crash_data$crash_date = 
  str_remove_all(crash_data$crash_date,"T[0-9]{2}:[0-9]{2}:[0-9]{2}.[0-9]{3}")
crash_data$crash_time = ymd_hm(paste(crash_data$crash_date,crash_data$crash_time),tz = "EST")
crash_data = crash_data[, -1]
```
#### Converting numeric type variables

```{r}
conv_func = function(x) as.double(x,digits=8) # conv_func converts to double
crash_data[,c(4,5)] =
  map_df(crash_data[,c(4,5)], conv_func)
crash_data[,c(3,6,7,8,9,10,11,12,13,16)] = # Numeric conversion
  map_df(crash_data[,c(3,6,7,8,9,10,11,12,13,16)], as.numeric)
```

## Visualization
In this part, plots will be conducted for New York city in general then for each borough


#### Contributing factor of crashes in New York city

 What are the contribuiting factors of motor vehicles crashes in New York city?
 Column plots have been plotted for top 10 contribuiting factors for vehicle 1 and 2.

```{r message=FALSE, warning=FALSE}
# vehicle 1
contrib1 = crash_data %>% 
  select(contributing_factor_vehicle_1) %>%
  group_by(contributing_factor_vehicle_1) %>%
  summarise(crash_count=n()) %>%
  mutate(rank = min_rank(desc(crash_count))) %>%
   filter(rank <= 10 & contributing_factor_vehicle_1!="NA") %>%
    arrange(rank)   # Top 10 contribution factor of crashes from vehicle 1

p1 =ggplot( contrib1, 
      aes(x=reorder(contributing_factor_vehicle_1,crash_count),crash_count)) +
    geom_col(fill="firebrick2") +ggtitle("Top 10 contributing factor vehicle 1") + theme(plot.title =   element_text(hjust = 0.5)) + xlab("contribuiting factor") + ylab("crashes count") + coord_flip()  # Plot of top 10 contribution factor of crashes from vehicle 1
```

```{r message=FALSE, warning=FALSE}
# vehicle 2
contrib2=crash_data %>% select(contributing_factor_vehicle_2)%>% 
  group_by(contributing_factor_vehicle_2) %>%
  summarise(crash_count=n()) %>%
  mutate(rank = min_rank(desc(crash_count))) %>%
   filter(rank <= 10 & contributing_factor_vehicle_2 !="NA") %>%
    arrange(rank)  # Top 10 contribution factor of crashes from vehicle 2

p2 =ggplot( contrib2, 
      aes(x=reorder(contributing_factor_vehicle_2,crash_count),crash_count)) +
    geom_col(fill="firebrick2") +ggtitle("Top 10 contributing factor vehicle 2") + theme(plot.title =   element_text(hjust = 0.5)) + xlab("contribuiting factor") + ylab("crashes count") + coord_flip()  # Plot of top 10 contribution factor of crashes from vehicle 2
```

#### Plot of contributing factors

```{r, fig.height=4}
grid.arrange(p1,p2)
```
The plots show that the first contributing factor from motor vehicles 1 and 2 is unspecified. Then the driver distraction


Monthly crashes data
The following three blocks of code aggreagte the data monthly, then conduct a plot of times series of total crashes from June 2012 to April 2021
```{r}
monthly_crash_raw=crash_data %>% 
  mutate(crash_year = year(crash_time), crash_month = month(crash_time)) %>%
  select(crash_year,crash_month,borough, number_of_persons_injured,number_of_persons_killed,number_of_pedestrians_injured,number_of_pedestrians_killed, number_of_cyclist_injured, number_of_cyclist_killed,number_of_motorist_injured,number_of_motorist_killed) %>%
  rename(pers_injured="number_of_persons_injured",pers_killed="number_of_persons_killed",pedest_injured="number_of_pedestrians_injured",pedest_killed="number_of_pedestrians_killed", cycl_injured="number_of_cyclist_injured",cycl_killed ="number_of_cyclist_killed",mot_injured="number_of_motorist_injured",mot_killed="number_of_motorist_killed") %>%
  group_by(crash_year,crash_month) # Monthly crashes data
```


```{r message=FALSE, warning=FALSE}
# aggregation by month for each variable
monthly_crash=monthly_crash_raw %>% select(-borough) %>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_"))) 
  monthly_crash = cbind(1:nrow(monthly_crash),monthly_crash) %>%
    rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
labels_x = monthly_crash$event_date
```


#### Monthly total crashes plot

```{r , fig.height=2 ,message=FALSE, warning=FALSE}
total_monthly_crash = 
  monthly_crash_raw %>% select(crash_year,crash_month) %>%
  summarise(total_event = n())
total_monthly_crash = cbind(total_monthly_crash,1:nrow(total_monthly_crash))%>% 
  rename(date_month ="...4")
P_mth_tot=ggplot(total_monthly_crash, aes(x=date_month,y=total_event,color="red3")) + geom_line() + ggtitle("Monthly total crashes plot") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("crashes count")  +
  theme(axis.text.x = element_text(angle=45), axis.ticks.length.x = unit(-.1, "cm"),legend.position = "none")
ggplotly(P_mth_tot, tooltip = c("total_event"))
```
There is in general a slow increase of total monthly crashes in New York city, with two abrupt decreases and low values in April 2016 and 2020, and 
heading to the lowest value in April 2021, last month of collection of the data.


#### Fatalities(Injured/killed) plot

Plots of monthly total injured and killed during motor vehicles collisions.

```{r, fig.height=2 }
monthly_crash_cat=monthly_crash %>% mutate(injured = pers_injured, killed=pers_killed) %>%
  select(date_vec,injured,killed) %>%
  pivot_longer(!date_vec, names_to = "fatalities", values_to = "monthly_count")
P_mth_crash=monthly_crash_cat %>%ggplot(aes(x=date_vec,y=monthly_count,group=fatalities,color=fatalities)) + geom_line() + ggtitle("Fatalities plot") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("crashes count")  +
  theme(axis.text.x = element_text(angle=45), axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank())
ggplotly(P_mth_crash, tooltip = c("fatality_category","monthly_count"))
```
The plots show that the injured plot has always been above of that of killed, suggesting that there have been in general les killed that injured during crashes.
Probably due to the low speed limit of New York city in genral. This need to be inverstigated thoroughly to reach a conclusion, because there might be some lurking variables.


#### Plot by fatalities categories
Here categories are: cyclist, motorist and pedestrians: injured and killed
```{r, fig.height=2 }
# Plot by variable 

monthly_crash_long = monthly_crash[,-c(2,3,4,5,12)]
monthly_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")

P_mth_long=monthly_crash_long %>% ggplot(aes(x=date_vec,y=monthly_count,color=fatality_category)) + geom_line() + ggtitle("Fatalities by category") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("crashes count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank())
ggplotly(P_mth_long, tooltip = c("fatality_category","monthly_count"))
```
The plots are in general: above that of motorists injured, followed by that of pedestrians injured, cyclists injured. There is no difference in general for the
number of motorists killed, pedestrians killed, and cyclists killed.


Monthly aggregation by borough

Aggregation is conducted for the five borough: Bronx, Brooklyn, Mahanttan, Queens and Staten Island.

BRONX
```{r message=FALSE, warning=FALSE}
bronx_crash=monthly_crash_raw %>% filter(borough=="BRONX") %>% select(-borough)%>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_")))
  bronx_crash = cbind(1:nrow(bronx_crash),bronx_crash) %>%
   rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
```

BROOKLYN
```{r message=FALSE, warning=FALSE}
brooklyn_crash=monthly_crash_raw %>% filter(borough=="BROOKLYN") %>% select(-borough)%>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_")))
  brooklyn_crash = cbind(1:nrow(brooklyn_crash),brooklyn_crash) %>%
    rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
```

MANHATTAN
```{r message=FALSE, warning=FALSE}
manhattan_crash=monthly_crash_raw %>% filter(borough=="MANHATTAN") %>% select(-borough)%>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_")))
  manhattan_crash = cbind(1:nrow(manhattan_crash),manhattan_crash) %>%
  rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
```

QUEENS
```{r message=FALSE, warning=FALSE}
queens_crash=monthly_crash_raw %>% filter(borough=="QUEENS") %>% select(-borough)%>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_")))
  queens_crash = cbind(1:nrow(queens_crash),queens_crash) %>%
   rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
```

STATEN ISLAND
```{r message=FALSE, warning=FALSE}
statenIsl_crash=monthly_crash_raw %>% filter(borough=="STATEN ISLAND") %>% select(-borough)%>%
  summarise(across(everything(), list(function(x) sum(x,na.rm = TRUE)))) %>%
  mutate(event_date=(paste(crash_year,crash_month,sep = "_")))
  statenIsl_crash = cbind(1:nrow(statenIsl_crash),statenIsl_crash) %>%
   rename(date_vec = "...1",pers_injured="pers_injured_1", pers_killed="pers_killed_1",
           pedest_injured="pedest_injured_1",pedest_killed="pedest_killed_1",
           cycl_injured="cycl_injured_1", cycl_killed="cycl_killed_1", mot_injured="mot_injured_1",mot_killed ="mot_killed_1")
```

#### Fatalities plot per borough
Plots by borough by categories
##### BRONX
```{r, fig.height=2 }
bronx_crash_long = bronx_crash[,-c(2,3,12)]
bronx_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")
 P_bronx = ggplot(bronx_crash_long, aes(x=date_vec,y=monthly_count,fill=fatality_category)) + geom_bar(position="stack", stat="identity") + ggtitle("Fatalities Bronx") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("fatality count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank()) + coord_cartesian(xlim=c(len_xaxis-24, len_xaxis))
 ggplotly(P_bronx, tooltip = c("fatality_category","monthly_count"))
```

##### BROOKLYN
```{r, fig.height=2 }
brooklyn_crash_long = brooklyn_crash[,-c(2,3,12)]
brooklyn_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")
 P_brooklyn = ggplot(brooklyn_crash_long, aes(x=date_vec,y=monthly_count,fill=fatality_category)) + geom_bar(position="stack", stat="identity") + ggtitle("Fatalities Brooklyn") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("fatality count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank()) + coord_cartesian(xlim=c(len_xaxis-24, len_xaxis))
 ggplotly(P_brooklyn,tooltip = c("fatality_category","monthly_count"))
```

##### MANHATTAN
```{r, fig.height=2 }
manhattan_crash_long = manhattan_crash[,-c(2,3,12)]
manhattan_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")
 P_manhattan = ggplot(manhattan_crash_long, aes(x=date_vec,y=monthly_count,fill=fatality_category)) + geom_bar(position="stack", stat="identity") + ggtitle("Fatalities Manhattan") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("fatality count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank()) + coord_cartesian(xlim=c(len_xaxis-24, len_xaxis))
 ggplotly(P_manhattan, tooltip = c("fatality_category","monthly_count"))
```

##### QUEENS
```{r, fig.height=2 }
queens_crash_long = queens_crash[,-c(2,3,12)]
queens_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")
 P_queens = ggplot(queens_crash_long, aes(x=date_vec,y=monthly_count,fill=fatality_category)) + geom_bar(position="stack", stat="identity") + ggtitle("Fatalities Queens") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("fatality count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.position="bottom",legend.title=element_blank()) + coord_cartesian(xlim=c(len_xaxis-24, len_xaxis))
 ggplotly(P_queens, tooltip = c("fatality_category","monthly_count"))
```

##### STATEN ISLAND
```{r, fig.height=2 }
statenIsl_crash_long = statenIsl_crash[,-c(2,3,12)]
statenIsl_crash_long %<>%
  pivot_longer(!date_vec, names_to = "fatality_category", values_to = "monthly_count")
 P_statenIsl = ggplot(statenIsl_crash_long, aes(x=date_vec,y=monthly_count,fill=fatality_category)) + geom_bar(position="stack", stat="identity") + ggtitle("Fatalities Staten Island") + 
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + scale_x_discrete(limits=labels_x,guide = guide_axis(check.overlap = TRUE))+
  ylab("fatality count")  +
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"), legend.position="bottom",legend.title=element_blank()) 
 ggplotly(P_statenIsl,tooltip = c("fatality_category","monthly_count"))
```

```{r}
boro_mot_injured = data.frame(bronx_crash$date_vec, bronx_crash$mot_injured,
                brooklyn_crash$mot_injured,manhattan_crash$mot_injured, 
                queens_crash$mot_injured,statenIsl_crash$mot_injured)
colnames(boro_mot_injured) = c("date_vec", "Bronx", "Brooklyn",
                            "Manhattan","Queens","StatenIsl")
```
Because motorists injured and pedestrians injured plots where above all the categories in general. There have been plot individually by categories as followed.

#### Motorists injured in NYC boroughs

```{r, fig.height=2 }
mot_injured_long = boro_mot_injured %>%
  pivot_longer(!date_vec, names_to = "boro", values_to = "inj_count")
P_mot_inj =ggplot(mot_injured_long, aes(x = date_vec, y=inj_count,color=boro)) + 
  geom_line() + ggtitle("Motorists injured in NYC boroughs") +
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + ylab("Motorists injured")  + 
  scale_x_discrete(limits=labels_x)+
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.title=element_blank()) 
 ggplotly(P_mot_inj,tooltip = c("inj_count","boro"))
```


```{r}
boro_peds_injured = data.frame(bronx_crash$date_vec, bronx_crash$pedest_injured,
                brooklyn_crash$pedest_injured,manhattan_crash$pedest_injured, 
                queens_crash$pedest_injured,statenIsl_crash$pedest_injured)
colnames(boro_peds_injured) = c("date_vec", "Bronx", "Brooklyn",
                            "Manhattan","Queens","StatenIsl")
```

#### Pedestrians injured in NYC boroughs


```{r, fig.height=2 }
peds_injured_long = boro_peds_injured %>%
  pivot_longer(!date_vec, names_to = "boro", values_to = "inj_count")
P_peds_inj =ggplot(peds_injured_long, aes(x = date_vec, y=inj_count,color=boro)) + 
  geom_line() + ggtitle("Pedestrians injured in NYC boroughs") +
  theme(plot.title =   element_text(hjust = 0.5))+  xlab("Date") + ylab("Motorists injured")  + 
  scale_x_discrete(limits=labels_x)+
  theme(axis.text.x = element_text(angle=45),axis.ticks.length.x = unit(-.1, "cm"),legend.title=element_blank()) 
 ggplotly(P_peds_inj,tooltip = c("inj_count","boro"))
```
Could we say from the last two plots that Brooklyn is riskier and staten island is safer for pedestrians and motorists? We need to be careful!!!! Statistical fallacies.
More studies need to be conducted to reach a conclusion.


This project can be improved by adding plots on map and more. Any suggestion will be greatly appreciated.
### Note: The file is large, updating software help limit issues
